Topological Equivariant Artist Model

### 2021-03-31 paper/paper.tex
currently submitted version

General purpose visualization libraries are often tasked with supporting many domains, leading the libraries to grow organically in ways that can lead to incoherent interfaces and brittle components. To adapt to modern library development needs, this paper presents a functional model of the how components provided by lower level libraries\cite{wongsuphasawatNavigatingWideWorld2021} can be guaranteed to be structure preserving maps from data to graphic. We target this level of library architecture because our aim is the libraries developers use to build automated and user facing tools, not those tools themselves; we are agnostic to what makes for a good visualization so long as the structure preserving constraints are satisfied. 
\begin{figure}[H]
  \centering 
  \subfloat[\label{fig:intro:continuity}]{%
    \includegraphics[width=1\columnwidth]{continuity.png}}\\
  \subfloat[\label{fig:intro:equivariance}]{%
    \includegraphics[width=1\columnwidth]{equivariance.png}}
\caption{In \autoref{fig:intro:continuity}, the line plot does not preserve \textit{continuity} because it implies that the discrete records are connected to each other, while the bar plot is \textit{continuity} preserving because it visually represents the records as independent data points. In \autoref{fig:intro:equivariance}, the data in blue is scaled by a factor of two, yielding the data in orange. To preserve \textit{equivariance}, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
\label{fig:intro:structure}
\end{figure}
The structure that components of a visualization library must preserve is \textit{continuity} and \textit{equivariance}, as illustrated in \autoref{fig:intro:structure}. \textit{Continuity} is the way in which the records in the dataset are connected to each other. In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval. \textit{Equivariance} means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side. In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from orange to blue.

Motivated by the challenge of re-architecturing the Python visualization library Matplotlib, we developed a functional model of library architecture, which we call the Topological Equivariant Artist Model (TEAM), to specify the constraints components must satisfy to be equivariant continuity preserving mappings from data to visual representation. A functional architecture derived from our model would gain a generalized data abstraction and well specified reusable components with minimal side effects\cite{loudenProgrammingLanguagesPrinciples2010}, in turn yielding improvements in maintainability, extendability, scaling, and support for concurrency. The contribution of this work is 
\begin{enumerate}
  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist}
  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code}
\end{enumerate}

\section{Related Work}
The notion that visualization is equivariant continuity preserving maps from data to visual representation is neither a new formalism nor a new implementation goal, but this work bridges the formalism and implementation in a functional manner at a building building blocks library level. When introducing the retinal variables, Bertin informally introduces the notion that continuity is preserved in the mark and  defines equivariance constraints (selective, associative, ordered, quantitative)\cite{bertinSemiologyGraphicsDiagrams2011a}. In the \textit{A Presentation Tool}(APT) model, Mackinlay embeds the continuity constraint in the choice of visualization type and generalizes the equivariance constraint to preserving a binary operator from one domain to another. The algebraic model of visualization proposed by Kindlmann and Scheidegger restrict equivarance to group actions, but explicitly define composable symmetric mappings from data space to representation space to graphic space. Our model targets monoid actions as they are more restrictive than all binary operations but more general than groups, and explicitly formalizes the way in which continuity is preserved as topologically maps. APT and the algebraic model are also targeted at visual designs and tools that automatically generate those designs, while our model targets the tools that build those tools.

Current visualization tools tend to either be architectured around a core data structure\cite{HeerSoftware2006} or explicitly implement the visual algorithm in terms of the data continuity the algorithm expects \cite{toryRethinkingVisualizationHighlevel2004}.  For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}. Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}. Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.  Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured. Our model facilitates unifying these APIs via an abstraction of data general enough to encompass most continuities, the clear separation of data, visual, and graphic transformations into functional components to mitigate side effects\cite{loudenProgrammingLanguagesPrinciples2010}, and by specifying the constraints these components must satisfy to map data to visualizations in a structure preserving manner.

\section{Topological Equivariant Artist Model}
\label{sec:math}
We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
from data $\mathscr{\dtotal}$ (\autoref{sec:math:data}) to graphic $\mathscr{\gtotal}$ (\autoref{sec:math:graphic}) fiber bundles. We decompose the artist into an indexing map from graphic to data (\autoref{sec:math:graphic:base}), a map from data components to visual components (\autoref{sec:math:artist:nu}), and map from visual components to graphic (\autoref{sec:math:artist:q}). In this section we discuss the formal properties of these fiber bundles and maps such that they can be used to specify an implementation, for example the one prototyped in \autoref{sec:code}.

@@

\subsubsection{Fiber Space: Variables}
\label{sec:math:data:fiber}
To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES,spivakDatabasesAreCategories2010} that binds the components of the fiber to variable names. Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset. He then defines the single variable set \fttype, which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname. The \fttype\ lookup is by name to specify that every component is distinct, since multiple components can have the same type \ftype. Given \fsection, the fiber for a one variable dataset is
\begin{equation}
    \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
\end{equation}
where \fsection\ is the schema binding variable name \fname\ to its datatype \ftype. A dataset with multiple variables has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
\begin{equation}
F = \ftotal_{\fsection(\fname_{1})}\times \ldots \ftotal_{\fsection(\fname_{i})} \ldots\times \ftotal_{\fsection(\fname_{n})}
\end{equation}
which is equivalent to 
\begin{equation}
    \label{eq:math:data:fiber:decompose}
    \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
\end{equation}
which allows us to decouple \dfiber\ into components $\dfiber_i$. Each component of \dfiber\ is a dimension of the topological fiber space and is specified by a tuple of the form (\fname, \ftype, \(\ftotal_{\fsection(\fname)}\)). 
\begin{figure}[htb]
  \centering 
  \includegraphics[width=\columnwidth]{fiber.png}
  \caption{These two datasets have the same base space \dbase. The plane is a representation of the fiber $\dfiber=\reals\times\reals$ for the variables (time, temperature) from \autoref{fig:intro:equivariance}, while the cube is the fiber $\reals\times\realsp\times\reals$ associated with (time, wind=(speed, direction))}
  \label{fig:math:data:fiber}
\end{figure}

In \autoref{fig:math:data:fiber} the plane fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{temperature}, \texttt{float}, \reals), while the cube fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{wind}, \texttt{wind}, \(\realsp\times\reals)\)) which encodes (\emph{speed}, \emph{direction}).

@@

A monoid \cite{Monoid2021} $\monoid$ is a set with an associative binary operator $\ast:\monoid \times \monoid\rightarrow \monoid$. A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties of associativity and identity. As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
\begin{equation}\label{eq:math:data:monoid:decompose}
\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
\end{equation}
of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$

Defining the monoid actions on the components serves as the basis for identifying the invariance\cite{kindlmannAlgebraicProcessVisualization2014} that must be preserved in the visual representation of the component. Monoids are commonly found in functional programming because a property of monoids is that the components can be composed into complex transformation pipelines \cite{yorgeyMonoidsThemeVariations}. For example, a component that converts \(\dfiber_i=time\) with \(\monoid_i\) to hours can be combined with a component that shifts \(time\) by 2 hours to yield \(hours_2\) with the same \(monoid_i\).

\subsubsection{Base Space: Continuity}
\label{sec:math:data:base}

 The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other. This is similar the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}, but our model treats keys as a pure reference to topology. Decoupling the keys from their semantics allows the metadata to be altered facilitates encoding of data where the independent variable may not be clear; for example the growth of a plant is dependent on both time and sunlight, and changing the changing the coordinate system or time resolution should have no effect on how the records are connected to each other.

@@

\subsubsection{Equivariant Topology: Graphic Base Space}
\label{sec:math:graphic:base}
\begin{figure}[htb]
  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
  \includegraphics[width=\columnwidth]{retraction_maps.png}
  \caption{The 0D scatter \dbasepoint\ and 1D line \dbasepoint\ are thickened into \gbase\ with coordinates $\gbasepoint=(\gx, \gy)$ that are a region in an idealized 2D screen. The image has the same dimension in \gbase\ as in \dbase.}
  \label{fig:math:graphic:base}
 \end{figure}

@@

\begin{table}[H]
  \renewcommand{\arraystretch}{2}
  \begin{tabulary}{\columnwidth}{|llL|}\hline
      \textbf{scale} & \textbf{group} & \textbf{constraint}\\ \hline
      nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
      ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
      interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
      ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\ \hline
  \end{tabulary}
  \label{tab:math:artist:nu}
\end{table}
We can state the conditions on \vchannel\ such that they cover the Stevens measurement scales\cite{stevensTheoryScalesMeasurement1946}, as defined in \autoref{tab:math:artist:nu}. This is because Stevens’ defined the measurement scales in terms of their mathematical group structure and a group is a monoid with inverses\cite{remlingAlgebraMath5353}. An example of monoid action equivariance is the preservation of partial ordering illustrated in \autoref{fig:math:artist:nu}.

@@

\section{Prototype}
\label{sec:code}
To build a prototype, we make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and exploit the Matplotlib transform stack to transform data coordinates into screen coordinates. While the artist is specified in a fully functional manner in \autoref{eq:math:artist:diagram}, we implement our prototype in a heavily object oriented manner. We do so mostly to more easily manage function inputs, especially parameters that are passed through to the structurally functional transform and draw methods.

@@

      def draw(self, renderer):
          # returns K, indexed on fiber then key 
          tau = self.E.view(self.axes) 
          # visual channel encoding applied fiberwise 
          mu = {p: nu(tau(c))
                   for p, (c, nu) in self.V.items()} 
          self.hat_q(**mu)
          # pass configurations off to the renderer
          super().draw(renderer)
  \end{minted}
 The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal. The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that can be used to query back into data to select a subset as described in \autoref{sec:math:data:sheaf}. To ensure the integrity of the section, \mintinline{python}{view} must be atomic, which means that the values cannot change after the method is called in draw until a new call in draw. We put this constraint on the return of the \mintinline{python}{view} method so that we do not risk race conditions.

>>>>

General purpose visualization libraries are often tasked with supporting many domains, leading the libraries to grow organically in ways that can lead to incoherent interfaces and brittle components. To adapt to modern library development needs, this paper presents a functional model of the how components provided by lower level libraries can be guaranteed to be structure preserving maps from data to graphic. We target this level of library architecture because our aim is the libraries developers use to build automated and user facing tools, not those tools themselves; our model is agnostic as to what makes for a good visualization so long as the structure preserving constraints are satisfied. 
\begin{figure}[H]
  \centering 
  \subfloat[\label{fig:intro:continuity}]{%
    \includegraphics[width=1\columnwidth]{continuity.png}}\\
  \subfloat[\label{fig:intro:equivariance}]{%
    \includegraphics[width=1\columnwidth]{equivariance.png}}
\caption{In \autoref{fig:intro:continuity}, the line plot does not preserve \textit{continuity} because it implies that the discrete records are connected to each other, while the bar plot is \textit{continuity} preserving because it visually represents the records as independent data points. In \autoref{fig:intro:equivariance}, the data in blue is scaled by a factor of two, yielding the data in orange. To preserve \textit{equivariance}, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
\label{fig:intro:structure}
\end{figure}
The structure that components of a visualization library must preserve is \textit{continuity} and \textit{equivariance}, as illustrated in \autoref{fig:intro:structure}. \textit{Continuity} is the way in which the records in the dataset are connected to each other and \textit{equivariance} means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side. In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval. In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from orange to blue.

To specify the constraints components must satisfy to be equivariant continuity preserving mappings from data to visual representation, we developed a functional model of library architecture, which we call the Topological Equivariant Artist Model (TEAM). Motivated by the challenge of re-architecturing the Python visualization library Matplotlib, a functional architecture derived from our model would gain a generalized data abstraction and well specified reusable components with minimal side effects\cite{loudenProgrammingLanguagesPrinciples2010}, in turn yielding improvements in maintainability, extendability, scaling, and support for concurrency. The contribution of this work is 
\begin{enumerate}
  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist}
  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code}
\end{enumerate}

\section{Related Work}
The notion that visualization is equivariant continuity preserving maps from data to visual representation is neither a new formalism nor a new implementation goal, but this work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library\cite{wongsuphasawatNavigatingWideWorld2021} level. When introducing the retinal variables, Bertin informally specifies that continuity is preserved in the mark and defines equivariance constraints in terms of data and visual variables being selective, associative, ordered, or quantitative\cite{bertinSemiologyGraphicsDiagrams2011a}. In the \textit{A Presentation Tool}(APT) model, Mackinlay embeds the continuity constraint in the choice of visualization type and generalizes the equivariance constraint to preserving a binary operator from one domain to another. The algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}, proposed by Kindlmann and Scheidegger, restricts equivariance to invertible transformations. Our model defines equivariance in terms of monoid actions, which are more restrictive than all binary operations but more general than invertible transformations. Their algebraic model also describes visualization as commutative mappings from data space to representation space to graphic space. Our model has the same notion of commutative mappings across analogous spaces, but uses topology to explicitly include continuity.

We emphasize continuity because visualization tools tend to either be architectured around a core data structure\cite{HeerSoftware2006} or explicitly implement the visual algorithm in terms of the data continuity the algorithm expects \cite{toryRethinkingVisualizationHighlevel2004}. For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}. Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}. Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.  Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured. Our model facilitates unifying these APIs via an abstraction of data general enough to encompass most continuities, the clear separation of data, visual, and graphic transformations into functional components to mitigate side effects, and by specifying the constraints these components must satisfy to map data to visualizations in a structure preserving manner.

\section{Topological Equivariant Artist Model}
\label{sec:math}
We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
from data $\mathscr{\dtotal}$ (\autoref{sec:math:data}) to graphic $\mathscr{\gtotal}$ (\autoref{sec:math:graphic}) fiber bundles. We decompose the artist into an indexing map from graphic to data (\autoref{sec:math:graphic:base}), a map from data components to visual components (\autoref{sec:math:artist:nu}), and map from visual components to graphic (\autoref{sec:math:artist:q}).

@@

\subsubsection{Fiber Space: Variables}
\label{sec:math:data:fiber}
To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES,spivakDatabasesAreCategories2010} that binds the components of the fiber to variable names. Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset. He then defines the single variable set \fttype, which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname. The \fttype\ lookup is by name to specify that every component is distinct, since multiple components can have the same type \ftype. Given \fsection, the fiber for a one variable dataset is
\begin{equation}
    \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
\end{equation}
where \fsection\ is the schema binding variable name \fname\ to its datatype \ftype. A dataset with multiple variables has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
\begin{equation}
F = \ftotal_{\fsection(\fname_{1})}\times \ldots \ftotal_{\fsection(\fname_{i})} \ldots\times \ftotal_{\fsection(\fname_{n})}
\end{equation}
which is equivalent to 
\begin{equation}
    \label{eq:math:data:fiber:decompose}
    \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
\end{equation}
which allows us to decouple \dfiber\ into components $\dfiber_i$. Each component of \dfiber\ is a dimension of the topological fiber space and is specified by a tuple of the form (\fname, \ftype, \(\ftotal_{\fsection(\fname)}\)). 
\begin{figure}[htb]
  \centering 
  \includegraphics[width=\columnwidth]{fiber.png}
  \caption{These two datasets have the same base space \dbase. The plane is a representation of the fiber $\dfiber=\reals\times\reals$ for the variables (time, temperature) from \autoref{fig:intro:equivariance}, while the cube is the fiber $\reals\times\reals\times\reals$ associated with (time, wind=(speed, direction))}
  \label{fig:math:data:fiber}
\end{figure}

In \autoref{fig:math:data:fiber} the plane fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{temperature}, \texttt{float}, \reals), while the cube fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{wind}, \texttt{wind}, \(\reals\times\reals\)) which encodes (\emph{speed}, \emph{direction}).

@@

A monoid \cite{Monoid2021} $\monoid$ is a set with an associative binary operator $\ast:\monoid \times \monoid\rightarrow \monoid$. A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties of associativity and identity. As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
\begin{equation}\label{eq:math:data:monoid:decompose}
\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
\end{equation}
of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber \((\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)\)

Defining the monoid actions on the components serves as the basis for identifying the invariance\cite{kindlmannAlgebraicProcessVisualization2014} that must be preserved in the visual representation of the component. Monoids are commonly found in functional programming because a property of monoids is that the components can be composed into complex transformation pipelines \cite{yorgeyMonoidsThemeVariations}.

\subsubsection{Base Space: Continuity}
\label{sec:math:data:base}
 The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other. This is similar the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}, but our model treats keys as a pure reference to topology. Decoupling the keys from their semantics allows the metadata to be altered facilitates encoding of data where the independent variable may not be clear; for example the growth of a plant is dependent on both time and sunlight, and changing the changing the coordinate system or time resolution should have no effect on how the records are connected to each other.

@@

\subsubsection{Equivalant Continuity: Graphic Base Space}
\label{sec:math:graphic:base}
\begin{figure}[htb]
  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
  \includegraphics[width=\columnwidth]{retraction_maps.png}
  \caption{The 0D scatter \dbasepoint\ and 1D line \dbasepoint\ are thickened into \gbase\ with coordinates $\gbasepoint=(\gx, \gy)$ that are a region in an idealized 2D screen. The image has the same dimension in \gbase\ as in \dbase.}
  \label{fig:math:graphic:base}
 \end{figure}

@@

\begin{table}[H]
  \renewcommand{\arraystretch}{2}
  \begin{tabulary}{\columnwidth}{|llL|}\hline
      \textbf{scale} & \textbf{group} & \textbf{constraint}\\ \hline
      nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
      ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
      interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
      ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\ \hline
  \end{tabulary}
  \caption{}
  \label{tab:math:artist:nu}
\end{table}
We can state the conditions on \vchannel\ such that they cover the Stevens measurement scales\cite{stevensTheoryScalesMeasurement1946}, as defined in \autoref{tab:math:artist:nu}. This is because Stevens’ defined the measurement scales in terms of their mathematical group structure and a group is a monoid with inverses\cite{remlingAlgebraMath5353}. An example of monoid action equivariance is the preservation of partial ordering illustrated in \autoref{fig:math:artist:nu}.

@@

\section{Prototype}
\label{sec:code}
To build a prototype, we make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and exploit the Matplotlib transform stack to convert data coordinates into screen coordinates. While the artist is specified in a fully functional manner in \autoref{eq:math:artist:diagram}, we implement our prototype in a heavily object oriented manner. We do so mostly to more easily manage function inputs, especially parameters that are passed through to the structurally functional transform and draw methods.

@@

      def draw(self, renderer):
          # returns K, indexed on fiber then key 
          tau = self.E.view(self.axes) 
          # visual channel encoding applied fiberwise 
          mu = {p: nu(tau(c))
                   for p, (c, nu) in self.V.items()} 
          self.hat_q(**mu)
          # pass configurations off to the renderer
          super().draw(renderer)
  \end{minted}
 The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal. The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that can be used to query back into data to select a subset as described in \autoref{sec:math:data:sheaf}. To ensure the integrity of the section, \mintinline{python}{view} must be atomic, which means that the values cannot change after the method is called in draw until a new call to draw. We put this constraint on the return of the \mintinline{python}{view} method so that we do not risk race conditions.
